---
title: "Introduction to the fastCRRC package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

`fastCRRC` is the R package to fit the marginal competing risks regression model for large correlated clustered observations. Let $i=1,2,\ldots,n$ denote the clusters where each cluster has $k=1, \ldots, M_i$ observations. For the $k$th member in cluster $i$,  let $T_{ik}$ and $C_{ik}$ denote the failure and right censoring times, respectively. Then for right censored data, the corresponding observable random variables become $X_{ik}= \min(T_{ik}, C_{ik})$ and $\xi_{ik} = I(T_{ik} \leq C_{ik})\epsilon_{ik}$ where $\epsilon_{ik} \in \{1, \ldots,l\}$ denotes the cause of failure,  $I(\cdot)$ is the indicator function, and $Z_{ik} = (Z_{1ik},...,Z_{pik})^{\top}$ is a $p \times 1$ vector of time-independent covariates for individual $k$ in cluster $i$. 

The marginal cumulative incidence function conditional on the covariates is defined as $F_1(t;Z_{ik}) = P(T_{ik} \leq t, \, \epsilon_{ik} = 1|Z_{ik})$ and the marginal subdistribution hazard is modeled as

\begin{equation}
\lambda_1(t; Z_{ik}) = \lambda_{10}(t) \exp(\beta'_{0}Z_{ik}),
\end{equation}

where $\lambda_{10}(\cdot)$ is a completely unspecified baseline subdistribution hazard function and $\beta_0$ is a $p \times 1$ vector of unknown regression parameters. 


## Usages

Installation of `fastCRRC` package can be done by
```{r results='hide', message=FALSE, warning=FALSE,eval=FALSE}
devtools::install_github(repo="edemprd/fastCRRC")
```

or
```{r results='hide', message=FALSE, warning=FALSE}
base::require("fastCRRC")
```


`fastCrrC()` function has the following arguments:
```{r eval=FALSE, message=FALSE, warning=FALSE}
fastCrrC(formula, data, subset,cause, id, fitter = c("fastCrr", "crrc"),B = 0, 
         bMeth = c("twoStep", "cluStep"), 
         mcontrols = control(gtol=1e-6,maxiter=1000, eps = 1E-6,getBreslowJumps = TRUE, 
                             standardize = TRUE,cindex = FALSE, multicore = FALSE, 
                             mcores=detectCores()-1))
```

See the detailed description from `help(fastCrrC)`.

## Data and model fit

We first simulate CCR data with two covariates, similar to the data simulation setting of Defor et al. (2024). 

```{r ,warning=FALSE}
library(fastCRRC)
set.seed(1)
dat <- simCR(n = 10, al = .3, ga = .6, clsize = 2:10, summary = TRUE)
```

There are two ways to fit the data to obtain the model estimates using the `fitter` argument in the `fastCrrC`. One uses the two-way linear scan with bootstrap for variance estimation (`fitter="fastCrr"`) and the other uses the traditional marginal CCR model using asymptotic variance estimation (`fitter="crrc"`). We recommend the two-way linear scan for $n > 5000$ or large sample or cluster and cluster sizes. We recommend that the number of bootstrap samples should be `B=100` when using the two way linear scan for large samples. However, for smaller sample sizes we recommend increasing the number of bootstrap samples to `B=200` or as desired to obtain stable standard errors. However, for either fitter arguments to print the standard errors $B$ must be $>0$.

```{r ,warning=FALSE,cache=TRUE}
# Two-way linear scan optimized for large datasets
fit1 <- fastCrrC(Surv(time, status) ~ Z.1 + Z.2, data = dat, B = 200, fitter = "fastCrr")
summary(fit1)

# Traditional CCR with asymptotic variance
fit2 <- fastCrrC(Surv(time, status) ~ Z.1 + Z.2, data = dat, B=100, fitter = "crrc")
summary(fit2)
```

#### Concordance index

If a C-index is desired, we have conveniently included its implementation following the procedures outlined by Wolbers et al. (2014) for CCR data. To compute and print the C-index we only allow it when `fitter="fastCrr"`.
```{r ,warning=FALSE,cache=TRUE}
fit1 <- fastCrrC(Surv(time, status) ~ Z.1 + Z.2, data = dat, B = 200, fitter = "fastCrr", mcontrols = control(cindex=TRUE))

summary(fit1)
```
Of note, the `mcontrols` arguments in `fastCrrC` introduces an array of options for the user aside setting the C-index. It also has specific arguments applicable to only when `fitter="crrc"`

#### Fitting large data

Next, we demonstrate how to optimize the model using the `fitter="fastCrr"` in the `fastCrrC` function when the cluster sizes are large. We can also take advantage of the `multicore` argument in  `mcontrols`.
```{r ,warning=FALSE,cache=TRUE}
## Generating the data
set.seed(1)
datL <- simCR(n = 200, al = .3, ga = .6, clsize = 2:500, summary = TRUE)
fitL <- fastCrrC(Surv(time, status) ~ Z.1 + Z.2, data = datL, B = 100, fitter = "fastCrr", 
                 mcontrols = control(cindex=TRUE, multicore = TRUE))
## Fitting the model
summary(fitL)

```

## Plotting the CIFs

To plot the CIFs we can easily use the `plot` function in our package. First, we can generate the data as follows using simCR(). However, for the purposes of illustration we generate five causes of failure.

```{r ,warning=FALSE,cache=TRUE,fig.height = 3, fig.width = 5, fig.align = "center",results="hide"}
## Generate the data using simCR() 
set.seed(1)
dat <- simCR(n = 100, al = .3, ga = .6, cen = .4, clsize = 2:5, summary = TRUE)
dat$status <- I(dat$status > 0)* sample(1:5, nrow(dat), TRUE)

## fit the model
foo <- fastCrrC(Surv(time, status) ~ Z.1 + Z.2, data = dat)

## plot CIF for the event of interest
plot(foo)
plot(foo, newdata = data.frame(Z.1 = .5, Z.2 = 1))
## Plotting CIFs that involves other causes will automatically update the fit
plot(foo, type = "all")
plot(foo, type = "all", newdata = data.frame(Z.1 = .5, Z.2 = 1))
## Stacking CIF
plot(foo, type = "stack")
plot(foo, type = "stack", newdata = data.frame(Z.1 = .5, Z.2 = 1))
```

## References
